---
title: "dust"
author: "C. Nell"
date: "7/23/2018"
output: 
  html_document:
    toc: TRUE
    toc_float: TRUE
    toc_depth: 4
---

```{r, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, tidy=TRUE,error = TRUE, eval = TRUE, message = FALSE, warning = FALSE, rows.print=5, cols.min.print=4, fig.width=6, fig.height=4.5)
```

```{r Load preliminary packages, include=FALSE}
library(dplyr) ## for data wrangling - %>% function
library(reshape2) ##melt and cast data
library(tidyr) # 'separate' function
library(readxl) #read xlsx files into r on mac computer
library(vegan) # dissimilarity matrix, permanova functions
library(tidyverse)
library(stringr)
library(ggplot2) # plotting
library(magrittr)
library(cowplot)
library(formatR)

```
### Logic   

**Background:** Dust entering the Sierras comes from both regional and global sources -- and the contribution of each source changes over time. There is evidence that microbial communities can differ based on dust source and timing of dust deflation, and that this can have subsequent effects on endophytic associations (Caroline Frank’s work) and soil microbial communities at the deposition location. We therefore wanted to know:  
  
###**Q1. Do dust-associated microbial communities entering the Sierra Nevada differ across space and time?**   

####Outputs:   
####1. Determine the number of new OTUs identified at each site per sampling event  
####2. NMDS of unweighted unifrac or jaccard dissimilarity. PERMANOVA with PERMDISP. 
*Limitation: our time comparisons may be skewed because we did not sterilize collectors in between sampling events. Time points may appear more similar to each other than in reality. It’s also important to run on presence-absence data only; we don’t want to capture community differences that are driven by post-depositional divergence in relative abundances.*  

```{r read in mapping data, include=FALSE}
map6<-read.csv("data/SierraMap6.csv", header = TRUE, 
               col.names =   c("SampleID","BarcodeSequence","LinkerPrimerSequence","Year","Month","SiteCode","RepNum","SiteRep","Site","Elevation","DateCode","DescName","Description"))%>%
  transform(Month=as.character(Month))%>% #transform to edit spellings
  mutate(Month = ifelse(Month == 'Oct','October', Month)) # consolidate multiple spellings of October

unique(map6$Month)

```  

#### Prepare data
Split taxonomy column into multiple variables
```{r}
# bacteria otu
B16S<-read.csv("data/filtered_table_w_metadata.csv", header = TRUE, stringsAsFactors = FALSE)

# split taxa groupings into new columns
source('dust_functions.R')
B16S<-split_taxa(B16S)
```  

```{r separate taxonomy, include=FALSE, eval=FALSE}
# extract the string between k__ and ; for the kingdoms
B16S$kingdom<-str_match(B16S$taxonomy, "k__(.*?);")[,2] #regex

# do the same for other taxa groupings 
B16S$phylum<-str_match(B16S$taxonomy, "p__(.*?);")[,2]
B16S$class<-str_match(B16S$taxonomy, "c__(.*?);")[,2]
B16S$order<-str_match(B16S$taxonomy, "o__(.*?);")[,2]
B16S$family<-str_match(B16S$taxonomy, "f__(.*?);")[,2]
B16S$genus<-str_match(B16S$taxonomy, "g__(.*?);")[,2]
B16S$species<-str_match(B16S$taxonomy, "s__(.*?)")[,2]

```  
  
#### Read in UniFrac distances    
UniFrac is a measure of B-diversity that uses phylogenetic information to compare community samples. Use with multivariate methods like principal coordinates (PCoA), NMDS, etc to explain differences among communities. Measured as the amount of evolutionary history that is unique to each group - the fraction of branch length in a phylogenetic tree associated with  ONLY the focal sample in pairwise comparison. With sequence data UniFrac values can be invfluencd by the number of sequences/sample and can use sequence jackknifing (Lozupone et al 2011. ISME).    
Use to compare phylogenetic lineages between groups, to cluster samples. 
```{r}
unifrac<-read.table('data/unweighted_unifrac_dm.txt')
unifrac_wt<-read.table('data/weighted_unifrac_dm.txt')

```  
Use heatmap to visualize sample dissimilarity based on UniFrac  
```{r, fig.width=10, fig.height=4, echo=FALSE}
## plot otu unifrac dissimilarities as heatmap
# first melt df so there is a new row for every pairwise combo
unifrac.melt<-unifrac%>%melt(variable.name='otu_1')%>%
  mutate(otu_2 = rep.int(colnames(unifrac), times=length(colnames(unifrac))))

# dark color indicates samples that were more similar to one another based on unifrac dissimilarity, blue = contains no similarities  
##reorder axis so grouped by more similar samples
heat.uni<-ggplot(data = unifrac.melt, aes(x = reorder(otu_1, value), y = reorder(otu_2, value)))+ 
  geom_tile(aes(fill = value))+
  scale_fill_gradient2(low = 'midnightblue', mid='deepskyblue3', high='yellow', midpoint = .5)+
  theme(axis.text.x = element_text(angle=90),axis.text = element_text(size=9))+
  labs(x='Sample', y='Sample', title='unweighted')

# make the same plot for weighted unifrac, compare
unifracwt.melt<-unifrac_wt%>%melt(variable.name='otu_1')%>%
  mutate(otu_2 = rep.int(colnames(unifrac_wt), times=length(colnames(unifrac_wt))))

heat.wt<-ggplot(data = unifracwt.melt, aes(x = reorder(otu_1, value), y = reorder(otu_2, value)))+ 
  geom_tile(aes(fill = value))+
  scale_fill_gradient2(low = 'midnightblue', mid='deepskyblue3', high='yellow', midpoint = .5)+
  theme(axis.text.x = element_text(angle=90),axis.text = element_text(size=9))+
  labs(x='Sample', y='Sample', title='weighted')

plot_grid(heat.uni, heat.wt, nrow=1, ncol=2)
## looks like weighting affects the similarity
# communities are more similar overall with the weighted unifrac - relative abundance 
# unweighted, most are 50% or more dissimilar. 
```  

```{r, include=FALSE, eval=FALSE}
str(map6) #grouping variables 
colnames(B16S) # community data

otu.taxa<-B16S%>%dplyr::select(OTU=X.OTU.ID, kingdom:species)
write.csv(otu.taxa, 'data/OTU_taxa_id.csv', row.names=FALSE) # save full # reorder columns data
```

```{r single and doubletons, eval=FALSE, include=FALSE}
otu.melt<-B16S%>%dplyr::select(-taxonomy:-species)%>%
  dplyr::select(OTU = X.OTU.ID, everything())%>%
  melt(id.vars=c('OTU'), variable.name='SampleID')

# remove singletons and doubletons
otu.melt$value_clean<-ifelse(otu.melt$value <= 2, NA, otu.melt$value)

# community df using cleaned data
otu.cast<-otu.melt%>%dcast(SampleID~OTU, value.var = 'value_clean')

## remove otus that were dropped with singletons and doubletons 
otu.mat<-otu.cast[,-1]
otu.new<-otu.mat[,colSums(otu.mat, na.rm=TRUE)> 0]
otu.new<-otu.new%>%mutate(SampleID=otu.cast$SampleID)%>%dplyr::select(SampleID, everything())

write.csv(otu.new, 'data/OTU_community_clean.csv', row.names=FALSE)
```

#### Prep data for multivariate anaylsis    
```{r read in cleaned data}
otu.taxa<-read.csv('data/OTU_taxa_id.csv') # otu ids
otu.comm<-read.csv('data/OTU_community_clean.csv')

```

```{r reorder data}
## the samples in the unifrac dist and community need to be in the same order
# need to order community based on unifraq

# make new df based on unifrac, matched to same order as unifrac
# match the order to colnames(unifrac), rewrite over same name
otu.comm<-otu.comm[match(colnames(unifrac), otu.comm$SampleID),]
# this says - for the community df, order the rows to match the order of the column names in unifrac, that match to SampleID

# reorder mapping data to community data 
grps<-map6[match(colnames(unifrac), map6$SampleID),]
# now that all the data (community, distances, grouping variables) are ordered the same, we can use permanova etc
```  

#### Permanova  
Permanova tests for differences in composition among groups  
Reminder - permanova is always based on pairwise distances/dissimilarities.  
```{r unifrac permanova}
set.seed(304)
#unifrac distances
str(grps) #look at data types to indetify error source
# Elevation is integer - change to factor 
grps$Elevation<-as.factor(grps$Elevation)
grps$Month<-as.factor(grps$Month)

ad.uni<-adonis2(unifrac~Month+Elevation, data=grps, permutations=1000, strata='Year')

ad.uni<-adonis2(unifrac~Month+Year+Elevation, data=grps, permutations=1000)
ad.uni

#unifrac distances with month nested in year
ad.uniNest<-adonis2(unifrac~Year/Month+Elevation, data=grps, permutations=1000)
ad.uniNest

#unifrac distances weighted by relative abundance
ad.uniwt<-adonis2(unifrac_wt~Month+Year+Elevation, data=grps, permutations=1000)
ad.uniwt
``` 
```{r other dist permanova, eval=FALSE, include=FALSE}
## are the results the same with other (non evolutionary) dissimiarlity indices?
dist.j<-vegdist(otu.comm[,-1], method='jaccard', na.rm=TRUE)
dist.bc<-vegdist(otu.comm[,-1], method='bray', na.rm=TRUE)

ad.bc<-adonis2(dist.bc~Month+Year+Elevation, data=grps, permutations=1000)
ad.bc

ad.j<-adonis2(dist.j~Month+Year+Elevation, data=grps, permutations=1000)
ad.j

```  
Month and elevation are significant in all tests. Year is not but that is probably because month is nested within year and that is not indicated in the model. 

**Manipulate code for test of interest** 

#### NMDS  
The NMDS often pairs well with the PERMANOVA because both are more robust to properties of community data. To gain proper fit may need to try multiple dissimilarity methods or tranformations.   
```{r configure nmds, echo=FALSE}
#metaMDS can take communty data or dissimilarities
mds.uni<-metaMDS(unifrac, k=2, autotransform=FALSE)
mds.uniwt<-metaMDS(unifrac_wt, k=2, autotransform=FALSE)

#stressplot(mds.uni)
#stressplot(mds.uniwt)

## both have good stress(below .15)
```
Plot ordination  
```{r nmds plot df}
# get NMDS scores for each SampleID
# scores are returned for each sampleid as a row, collapses dissimilarity data into 2 dimensions
nmds.uni<-data.frame(scores(mds.uni))
nmds.uni<-nmds.uni[match(grps$SampleID,rownames(nmds.uni)),]%>%cbind(grps) # match the rows and bind with the grouping data for plotting

# plotting data for nmds for weighted unifrac
nmds.uniwt<-data.frame(scores(mds.uniwt))
nmds.uniwt<-nmds.uniwt[match(grps$SampleID,rownames(nmds.uniwt)),]%>%cbind(grps) # reorder rows and bind to grouping data
```
```{r plot nmds}
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, shape=Month))
```
**Manipulate plot code to make figure as desired**  

```{r plot nmds with Year test 1}
#Not working to assign Year as a grouping factor
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(size=Year, shape=Month))

# converting variables between numeric and factors changes how ggplot will plot them if hitting errors
str(nmds.uni$Year)
nmds.uni<-nmds.uni%>%transform(Year = as.factor(Year))

ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(size=Year, shape=Month))
```

```{r plot nmds with Year test 2}
#Not working to assign Year as a grouping factor
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(size=Month, color=Year))

# size for discrete scale is a little funky
# also there are 2 different levels for the month of october due to spelling differences - correction added upstream to the grouping dataframe
```



```{r}
#Not running
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, shape=Site), size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)

# to use scale_color_gradient2, the assigned variable must be continuous (as.numeric) 
# a shape scale is most useful for categorical data  
# data type can be: numeric, factor, integer, and character. character and factor are very similar but are treated slightly differently

str(nmds.uni$Elevation)
# to convert to numeric back from factor you have to be careful
# going directly factor to numeric does not work. to preserve the values in a factor, convert to character, then numeric
nmds.uni$Elevation<-as.character(nmds.uni$Elevation)
nmds.uni$Elevation<-as.numeric(nmds.uni$Elevation)

ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, shape=Site), size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)

```
Separates well by elevation/site, within each elevation the months are grouped together for the most part.  


```{r}
ggplot(nmds.uniwt, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Year, shape=Month), size=4)+
  scale_color_gradient2(low='turquoise', high='red', midpoint=1500)
  #Not running - you need to change Year to categorical using as.factor() 
```
```{r}
## edit shape scale - the + is hard to see
# https://ggplot2.tidyverse.org/reference/scale_shape.html

# Show a list of available shapes
df_shapes <- data.frame(shape = 0:25)
ggplot(df_shapes, aes(0, 0, shape = shape)) +
  geom_point(aes(shape = shape), size = 5, fill = 'red') +
  scale_shape_identity() +
  facet_wrap(~shape) +
  theme_void()

# shapes in red let you manipulate the color & fill (rather than just color)
# color then corresponds to the shape outline, and fill to the inside
# let's use the fill to indicate the Year, shape for the Month, and color for Elevation
```
```{r editing plots}
# let's use the fill to indicate the Year, shape for the Month

# color = Elevation
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation), size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  theme_bw()

# fill = Year
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, fill=Year), size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  theme_bw()
# note - this is the same as before even with fill = Year added because we are using a shape that cannot be filled

# fill = Year, shape = 21
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, fill=Year), shape = 21,color='black', size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  theme_bw()
# shape = 21 is the filled circle

# shape = Month
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, fill=Year, shape=Month),color='black', size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  scale_shape_manual(values=c(21, 22, 24))+ # the values correspond to the shapes above
  theme_bw()

# and to change the color of the fill:
ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, fill=Year), shape = 21,color='black', size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  scale_shape_manual(values=c(21, 22, 24))+ # the values correspond to the shapes above
  scale_fill_manual(values=c('black','white'))+
  theme_bw()
```
```{r}
# lets add the stress to the plot
mds.uni$stress #the stress

ggplot(nmds.uni, aes(NMDS1, NMDS2))+
  geom_point(aes(color=Elevation, fill=Year, shape=Month),color='black', size=3)+
  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)+
  scale_shape_manual(values=c(21, 22, 24))+ # the values correspond to the shapes above
  scale_fill_manual(values=c('black','white'))+
  theme_bw()+
  annotate('text', x=0.3, -0.25, label=paste0('Stress = ',round(mds.uni$stress,2)))


```
  

####NMDS vs PCoA  
NMDS is an unconstrained ordination method to visualize multivariate data in fewer dimensions. Depending on the properties of your data and your questions, different methods of ordination may be appropriate. PCoA (or MDS) is a metric version of NMDS, meaning that PCoA is a Euclidean representation of dissimilarities. So, like NMDS PCoA uses dissimilarities, but it is euclidean rather than rank-order which is used to preserve distances hierarchically. NMDS and distance-based ordinations/tests are often easier to use with community data because 0's and 1's, rare species are common and do not configure in euclidean space. 

In general, ordination & multivaraite methods are defined by (1) whether or not using dissimilarity, (2) whether or not euclidean, (3) test for differences vs explaining differences  ...  

PCoA, like PCA, returns a set of orthogonal axes with associated eignevalues that measure their importance. PCoA and PCA should produce the same results if the PCA is calculated on a covariance matrix with scaling of 1. 

```{r include=FALSE, eval=FALSE}
#PCoA - unconstrained, euclidean
pcoa.uni<-capscale(unifrac~1,data=grps)
head(summary(pcoa.uni))

pcoa.df<-data.frame(scores(pcoa.uni, display='sites'))
pcoa.df# site scores correspond to the SampleID

# bind with grouping data, in order
pcoa.df<-pcoa.df[match(grps$SampleID,rownames(pcoa.df)),]%>%cbind(grps)


```  
**plot the pcoa** 

```{r include=FALSE, eval=FALSE}
##dbrda
#pcoa.uni<-capscale(unifrac~Elevation+Month,data=grps)

#not working
#ggplot(pcoa.uni, aes(NMDS1, NMDS2))+
#  geom_point(aes(color=Elevation, shape=Site), size=3)+
#  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)


#ggplot(pcoa.df, aes(NMDS1, NMDS2))+
#  geom_point(aes(color=Elevation, shape=Site), size=4)+
#  scale_color_gradient2(low='turquoise', mid='yellow', high='red', midpoint=1500)

```




###**Q2. If differences exist, are there particular taxa that contribute disproportionately to those differences?**    
*Note: Can’t run any analyses that rely on relative abundance data (e.g., simper), Address using absence-presence data*   
  
#### Outputs:  

####1. Indicator species analysis to identify taxon-habitat association patterns.  
This procedure identifies OTUs as indicator species independently from their abundance in the total data set. 
####Steps (from https://www.nature.com/articles/ismej2015238):   

  a. single- and doubleton OTUs are removed as they hold little indicator informationusing the multipatt function (number of permutations=9999) implemented in the indicspecies R package (De Cáceres et al., 2010). 
  b. To account for multiple testing, P-values were corrected by calculating false discovery rates (q-values) with the q-value function implemented in the BiocLite R package (Dabney and Storney, 2014). Indicator OTUs with q<0.05 were considered significant. 
  c. Indicator taxa were represented in bipartite networks by using the edge-weighted spring embedded algorithm layout implemented in Cytoscape v.3.0.2 (Shannon et al., 2003) where point biserial correlation values, a measure of species–habitat association, were used to weight the edges between nodes constituting the habitats and indicator OTUs (Hartmann et al., 2015). 
  d. We further mapped these indicator OTUs on taxonomic networks generated in Cytoscape v.3.0.2 to investigate potential taxa–habitat associations (Hartmann et al., 2015). I
  e. Indicator OTUs classified at the genus level were displayed in a taxonomic tree generated in iTOL (Letunic and Bork, 2011) together with the positive point biserial correlation values associated with each habitat.  
  f. To find patterns of co-occurrence among OTUs characteristic of the habitats studied, we analysed correlations among the relative abundances of all indicator OTUs (co-correlations) by calculating Spearman rank correlation values (Spearman, 1904) in R with the function corr.test implemented in the package psych. Multiple testing was accounted for by correcting P-values with a false discovery rates of 5% (q-value<0.05).  
  g. Bacterial and fungal indicator OTUs that were significantly co-correlated were displayed in networks with the edge-weighted spring embedded algorithm layout implemented in Cytoscape where Spearman correlation values were used to weight the edges between the nodes representing the OTUs (Hartmann et al., 2015).”   


###**Q3. Can the observed variation in XXX be explained by provenance? By nutrient composition?**
####Outputs: 
####1. Multiple regression on distance matrix  

